Historical aerial surveys map long-term changes of forest cover and structure in the central Congo Basin

### 2020-02-09 manuscript/main.Rmd
tilde

abstract: |
   Given the impact of tropical forest disturbances on atmospheric carbon emissions, biodiversity and ecosystem productivity, accurate long term reporting of Land-Use and Land-Cover (LULC) change in the pre-satellite era (< 1972) is an imperative. Here, we used a combination of historical (1958) aerial photography and contemporary remote sensing data to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi (DR Congo), in the central Congo Basin. Our study leveraged Structure-from-Motion and a Convolutional Neural Network based LULC classifier, using synthetic landscapes based image augmentation, to map historical forest cover across a large orthomosaic (93 431 ha) geo-referenced to ~4.7 $\pm$ 4.3 m at sub meter resolution. A comparison with contemporary LULC data showed a shift from previously highly regular industrial deforestation of large areas, to discrete smallholder farming clearing, increasing landscape fragmentation but also providing opportunties for substantial forest regrowth. We estimated above ground carbon gains through reforestation to range from 811 to 1592 Gg C, partially offseting historical deforestation (2416 Gg C), in our study area. Efforts to quantify long term canopy texture changes and their link to above ground carbon had limited to no success. Our analysis provides methods and insights into key spatial and temporal patterns of deforestation and reforestation at a multi-decadal scale, providing a historical context for past and ongoing forest research in the area.
keywords: |
  Aerial survey, data recovery, CNN, Deep Learning, SfM, Congo Basin
acknowledgement: |
  This research was supported through the Belgian Science Policy office COBECORE project (BELSPO; grant BR/175/A3/COBECORE). K.H. acknowledges funding from the European Union Marie Skłodowska-Curie Action (project number 797668). K.H. grateful for the support of T.d.H. volunteering his time in scanning the images.
authorcontributions: |
  K.H. conceived and designed the study, analyzed the data, prepared figures, tables and authored the final draft of the manuscript. T.d.H. scanned all image data. E.K. and T.d.H. provided plot based AGC estimates. T.D., K.J., E.K, H.B., P.S., F.V.,S.M., M.A., J.V.D.B., H.V and L.W. reviewed the final manuscript.
conflictsofinterest: |
  The authors declare no conflict of interest. The founding sponsors had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript, an in the decision to publish the results.
output: rticles::mdpi_article
---

@@

Our analysis provides a first spatially explicity historical composite of aerial survey images in support of mapping land-use and land-cover within the Congo Basin. The use of high resolution historical images combined with SfM image processing techniques allowed us to mosaic old imagery across a large extent. The final orthomosaic composition of the Yangambi region provided an image scene covering approximately 733 million pixels across 93 431 ha with a resolution of 0.88 m / pixel (or ~23x36 km, Figure 2). The overall spatial accuracy of the SfM orthomosaic composition using the sparse cloud DEM (with a resolution of 45.8 m/pixel) was limited to approximately 23 m. Further georeferencing outside the SfM workflow reduced the mean error at the ground control points to 5.3 $\pm$ 4.9 px (~4.7 $\pm$ 4.3 m), with a median error of 2.9 px (2.6 m). The orthomosaic served as input for all subsequent LULCC analysis with all derived maps provided with the manuscript repository (see data & code availability statements below).

>>>>

abstract: |
   Given the impact of tropical forest disturbances on atmospheric carbon emissions, biodiversity and ecosystem productivity, accurate long term reporting of Land-Use and Land-Cover (LULC) change in the pre-satellite era (< 1972) is an imperative. Here, we used a combination of historical (1958) aerial photography and contemporary remote sensing data to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi (DR Congo), in the central Congo Basin. Our study leveraged Structure-from-Motion and a Convolutional Neural Network based LULC classifier, using synthetic landscapes based image augmentation, to map historical forest cover across a large orthomosaic (~93 431 ha) geo-referenced to ~4.7 $\pm$ 4.3 m at sub meter resolution. A comparison with contemporary LULC data showed a shift from previously highly regular industrial deforestation of large areas, to discrete smallholder farming clearing, increasing landscape fragmentation but also providing opportunties for substantial forest regrowth. We estimated above ground carbon gains through reforestation to range from 811 to 1592 Gg C, partially offseting historical deforestation (2416 Gg C), in our study area. Efforts to quantify long term canopy texture changes and their link to above ground carbon had limited to no success. Our analysis provides methods and insights into key spatial and temporal patterns of deforestation and reforestation at a multi-decadal scale, providing a historical context for past and ongoing forest research in the area.
keywords: |
  Aerial survey, data recovery, CNN, Deep Learning, SfM, Congo Basin
acknowledgement: |
  This research was supported through the Belgian Science Policy office COBECORE project (BELSPO; grant BR/175/A3/COBECORE). K.H. acknowledges funding from the European Union Marie Skłodowska-Curie Action (project number 797668). K.H. grateful for the support of T.d.H. volunteering his time in scanning the images.
authorcontributions: |
  K.H. conceived and designed the study, analyzed the data, prepared figures, tables and authored the final draft of the manuscript. T.d.H. scanned all image data. E.K. and T.d.H. provided plot based AGC estimates. T.D., K.J., E.K, H.B., P.S., F.V.,S.M., M.A., J.V.D.B., H.V and L.W. reviewed the final manuscript.
conflictsofinterest: |
  The authors declare no conflict of interest. The founding sponsors had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript, an in the decision to publish the results.
output: rticles::mdpi_article
---

@@

Our analysis provides a first spatially explicity historical composite of aerial survey images in support of mapping land-use and land-cover within the Congo Basin. The use of high resolution historical images combined with SfM image processing techniques allowed us to mosaic old imagery across a large extent. The final orthomosaic composition of the Yangambi region provided an image scene covering approximately 733 million pixels across ~93 431 ha with a resolution of 0.88 m / pixel (or ~23x36 km, Figure 2). The overall spatial accuracy of the SfM orthomosaic composition using the sparse cloud DEM (with a resolution of 45.8 m/pixel) was limited to approximately 23 m. Further georeferencing outside the SfM workflow reduced the mean error at the ground control points to 5.3 $\pm$ 4.9 px (~4.7 $\pm$ 4.3 m), with a median error of 2.9 px (2.6 m). The orthomosaic served as input for all subsequent LULCC analysis with all derived maps provided with the manuscript repository (see data & code availability statements below).

### 2020-02-09 manuscript/main.tex
tilde

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{Given the impact of tropical forest disturbances on atmospheric carbon
emissions, biodiversity and ecosystem productivity, accurate long term
reporting of Land-Use and Land-Cover (LULC) change in the pre-satellite
era (\textless{} 1972) is an imperative. Here, we used a combination of
historical (1958) aerial photography and contemporary remote sensing
data to map long-term changes in the extent and structure of the
tropical forest surrounding Yangambi (DR Congo), in the central Congo
Basin. Our study leveraged Structure-from-Motion and a Convolutional
Neural Network based LULC classifier, using synthetic landscapes based
image augmentation, to map historical forest cover across a large
orthomosaic (93 431 ha) geo-referenced to \textasciitilde{}4.7 \(\pm\)
4.3 m at sub meter resolution. A comparison with contemporary LULC data
showed a shift from previously highly regular industrial deforestation
of large areas, to discrete smallholder farming clearing, increasing
landscape fragmentation but also providing opportunties for substantial
forest regrowth. We estimated above ground carbon gains through
reforestation to range from 811 to 1592 Gg C, partially offseting
historical deforestation (2416 Gg C), in our study area. Efforts to
quantify long term canopy texture changes and their link to above ground
carbon had limited to no success. Our analysis provides methods and
insights into key spatial and temporal patterns of deforestation and
reforestation at a multi-decadal scale, providing a historical context
for past and ongoing forest research in the area.}

@@

Our analysis provides a first spatially explicity historical composite
of aerial survey images in support of mapping land-use and land-cover
within the Congo Basin. The use of high resolution historical images
combined with SfM image processing techniques allowed us to mosaic old
imagery across a large extent. The final orthomosaic composition of the
Yangambi region provided an image scene covering approximately 733
million pixels across 93 431 ha with a resolution of 0.88 m / pixel (or
\textasciitilde{}23x36 km, Figure 2). The overall spatial accuracy of
the SfM orthomosaic composition using the sparse cloud DEM (with a
resolution of 45.8 m/pixel) was limited to approximately 23 m. Further
georeferencing outside the SfM workflow reduced the mean error at the
ground control points to 5.3 \(\pm\) 4.9 px (\textasciitilde{}4.7
\(\pm\) 4.3 m), with a median error of 2.9 px (2.6 m). The orthomosaic
served as input for all subsequent LULCC analysis with all derived maps
provided with the manuscript repository (see data \& code availability
statements below).

>>>>

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{Given the impact of tropical forest disturbances on atmospheric carbon
emissions, biodiversity and ecosystem productivity, accurate long term
reporting of Land-Use and Land-Cover (LULC) change in the pre-satellite
era (\textless{} 1972) is an imperative. Here, we used a combination of
historical (1958) aerial photography and contemporary remote sensing
data to map long-term changes in the extent and structure of the
tropical forest surrounding Yangambi (DR Congo), in the central Congo
Basin. Our study leveraged Structure-from-Motion and a Convolutional
Neural Network based LULC classifier, using synthetic landscapes based
image augmentation, to map historical forest cover across a large
orthomosaic (\textasciitilde{}93 431 ha) geo-referenced to
\textasciitilde{}4.7 \(\pm\) 4.3 m at sub meter resolution. A comparison
with contemporary LULC data showed a shift from previously highly
regular industrial deforestation of large areas, to discrete smallholder
farming clearing, increasing landscape fragmentation but also providing
opportunties for substantial forest regrowth. We estimated above ground
carbon gains through reforestation to range from 811 to 1592 Gg C,
partially offseting historical deforestation (2416 Gg C), in our study
area. Efforts to quantify long term canopy texture changes and their
link to above ground carbon had limited to no success. Our analysis
provides methods and insights into key spatial and temporal patterns of
deforestation and reforestation at a multi-decadal scale, providing a
historical context for past and ongoing forest research in the area.}

@@

Our analysis provides a first spatially explicity historical composite
of aerial survey images in support of mapping land-use and land-cover
within the Congo Basin. The use of high resolution historical images
combined with SfM image processing techniques allowed us to mosaic old
imagery across a large extent. The final orthomosaic composition of the
Yangambi region provided an image scene covering approximately 733
million pixels across \textasciitilde{}93 431 ha with a resolution of
0.88 m / pixel (or \textasciitilde{}23x36 km, Figure 2). The overall
spatial accuracy of the SfM orthomosaic composition using the sparse
cloud DEM (with a resolution of 45.8 m/pixel) was limited to
approximately 23 m. Further georeferencing outside the SfM workflow
reduced the mean error at the ground control points to 5.3 \(\pm\) 4.9
px (\textasciitilde{}4.7 \(\pm\) 4.3 m), with a median error of 2.9 px
(2.6 m). The orthomosaic served as input for all subsequent LULCC
analysis with all derived maps provided with the manuscript repository
(see data \& code availability statements below).
